Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
fix viterbi equations

\begin{figure*}
\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}}\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}}\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}}\\
V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}}\\
V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}}\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}}\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}}\\
V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}}\\
V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\\[\gum]
\multicolumn3{l}{%
a'_{s} = - \log a_{s} 
\qquad
e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}
\qquad
V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
But like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Because the cases for all three states are so similar, 
we~show only the code for the match state:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{smallverbatim}
viterbi' :: State -> Int -> Int -> (Score, [State])
viterbi' Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
                                          , trans i_m Ins
                                          , trans d_m Del 
                                          ]
                  , Mat : path )
  where (prevV, path) = viterbi prev (j-1) (i-1)
        trans t prev = prevV + tProb hmm (j-1) t
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

MRFy treats the beta 
strands, which are build into the model,
 as ``beads'' which can be slid along the query sequence to determine
 a \emph{placement} of the beta strands.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of any subsequence that appears
\emph{between} two beta strands can be computed exactly and
efficiently using Viterbi's algorithm.
(The result of the exact computation is a probability that is
\emph{conditioned} on the placement.)
MRFy searches for placements stochastically,
trying to maximize the product of these probabilities.
We~are exploring several
heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.
Our~exploration of these heuristics relies on higher-order functions.

We wrote a basic search function in only twenty-two lines of code.
The central logic of this search function is simply:
\begin{smallverbatim}
search' (s1:s2:seeds) oldPop hist age =
  let newPop = mutate oldPop
      score = fst $ minimum newPop
      newHist = score : hist
    in if accept newHist age then
          if terminate newHist age then
            (minimum newPop, newHist)
          else
            search' seeds newPop newHist (age + 1)
       else
         if terminate hist age then
           (minimum oldPop, hist)
         else
           search' seeds oldPop hist (age + 1)
\end{smallverbatim}
As one of the parameters to the top-level \texttt{search} function,
we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques.
We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies.
For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions.
In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}.
We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.
In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this.
We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.

We note that \texttt{seed} in \texttt{search} comes from a monadic random 
number generator in Main, but we wished for search itself to be pure at this 
level.
One advantage is that this allows for testing of the \texttt{search} 
function with no randomness at all: if we pass in a known number as the seed, 
behavior is entirely deterministic.

>>>>

\begin{figure*}
\newcommand\vsum[2]{#2&{}+{}& #1}

\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
\vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
\vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
\end{array} \right.\\[\gum]
\multicolumn3{l}{%
a'_{s} = - \log a_{s} 
\qquad
e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}
\qquad
V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
But like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Because the cases for all three states are so similar, 
we~show only the code for the match state:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{viterbi}
\begin{smallverbatim}
viterbi' :: State -> Int -> Int -> (Score, [State])
viterbi' Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
                                          , trans i_m Ins
                                          , trans d_m Del 
                                          ]
                  , Mat : path )
  where (prevV, path) = viterbi prev (j-1) (i-1)
        trans t prev = prevV + tProb hmm (j-1) t
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

MRFy treats the beta 
strands, which are build into the model,
 as ``beads'' which can be slid along the query sequence to determine
 a \emph{placement} of the beta strands.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of any subsequence that appears
\emph{between} two beta strands can be computed exactly and
efficiently using Viterbi's algorithm.
(The result of the exact computation is a probability that is
\emph{conditioned} on the placement.)
The likelihood is computed by a function \texttt{score},
which given a model and a {placement} returns the {placement}
paired with a \emph{score}.
A~score is a negated log probability, so lower scores are better.

MRFy searches for placements stochastically,
trying to minimize scores.
We~are exploring several stochastic algorithms: random-mutation hill
climbing, simulated  
annealing, multistart simulated annealing, and a genetic algorithm.
\remark{It would be pleasant to connect these to the functions below,
but we probably lack room.}
Our~exploration relies on higher-order functions.

MRFy's search works with a sequence of \emph{generations}.
A~generation is a collection of scored \emph{placements}.
We~pass the
search function a \texttt{Scorer}, which is the result of partially
applying \texttt{score} to the model.
\smallverbatiminput{scoredecl}
A~search algorithm is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallverbatiminput{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may not be any better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} considers the score of the best
placement of the trial generation, along with the history of previous
best scores.
It~also considers the \texttt{Age} of the trial generation, which is
the total number of 
generations tried.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the history of scores and the age of the current
generation. 
\end{itemize}
This interface allows for considerable variation.
For example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The interface \texttt{quit} makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.

We hope you find the code simple and straightforward:
\smallverbatiminput{search}

### 2012-03-10 search.tex
fix viterbi equations

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Scored Age])
search strat score (s:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s
  runFrom :: [Seed] -> [Scored Placement] -> [Scored Age]
          -> Age -> (Scored Placement, [Scored Age])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist =
          Scored age (scoreOf $ minimum newPop) : oldHist
        (newPop, newHist) =
          if accept strat s2 newHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)

>>>>

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Score])
search strat score (s:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s
  runFrom :: [Seed] -> [Scored Placement] -> [Score]
          -> Age -> (Scored Placement, [Score])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist = (scoreOf $ minimum newPop) : oldHist
        (newPop, newHist) =
          if accept strat s2 newHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)
